Physics-based protein structure prediction and design using the confinement method

### 2013-01-09 Paper/paper.tex
Al. Done with Chamaleon sequences

As a first step, we performed several control experiments to verify that our implementation of the
confinement method produces results compatible with previous calculations reported in the
literature.

The method has previously been applied to a 16 amino acid residue $\beta$-hairpin from
protein G, known as BHP \cite{Cecchini2009}. We calculated the free energy difference between two different
conformations of the peptide: (1) the native conformation, called bhp1, with a two-stranded
$\beta$-sheet; and (2) a conformation, called bhp3, which has a three-stranded $\beta$-sheet.
Analysis of long (4 $\mu$s) equilibrium simulations \cite{Cecchini2009},\cite{Krivov2004}  shows that bhp1 is the more favorable
configuration by 1.8 kcal/mol. Using the confinement method, we obtain a value 1.7 kcal/mol, which
is in good agreement with the equilibrium simulations and with previous calculations using the
confinement method \cite{Cecchini2009} (see Supporting Information for further details).

Previous applications of the confiment method have focused on relatively short peptides, up to $17$
residues in length. In this work we apply the confinement method to
larger proteins. On that direction, We first test the confinement method for a chamelon sequence and 
found that it can correctly pick out the prefential structure of the chamelon sequence. Other 
applications we explore in the present work is the re-scoring, or
metaprediction, of structures submitted during the Critical Assessment of Structure Prediction
(CASP) experiment (described in detail later). We computed the relative free energies of predictions
submitted during CASP9, with the expectation that the most native-like prediction will have
the lowest free energy. As a control, for several targets we also calculated the relative free
energy of the experimentally determined structure. As
Table~\ref{table:casp_control} shows, in most cases, the confinement method correctly assigns a
lower free energy to the experimentally determined structure than to any of the decoys. 
Most interestingly, we decomposed the free energy into its per residue component, which help 
us to identify the residues which stabilize or destabilize a particular conformation of the protein.
%Although differentiating between experimentally determined structures and computer generated predictions is
%not a stringent test of a scoring method---for example, examining the residue-residue packing can
%easily distinguish between the two [ref Rosetta Holes]---the results none the less serve as a useful
%"sanity check".

@@

In general, proteins with similar sequences tend to have similar structures. This idea is the basis
of comparative modeling and fold recognition in protein structure prediction. There are, however,
examples---often referred to as chameleon sequences---of proteins with similar sequences that have
remarkably different structures. Orban and co-workers have designed a series of 56-residue proteins
(based on Protein G) that adopt one of two different folds depending on small changes in sequence
(see Figure~\ref{fig:orban}). Sequences that adopt a mixed alpha/beta structure similar to Protein G
are denoted as ``GB'', while sequences that form a three-helix bundle are denoted as ``GA''. One
pair of sequences (GA88/GB88) are 88 percent identical in sequence and differ in seven positions. Another pair
(GA95/GA95) are 95 percent identical and differ in three positions. The last pair (GA98/GB98) differ
only in a single tyrosine to alanine mutation.

@@

To avoid this potential problem, we instead computed the relative free energy two different
structural models for each sequence. One model is based on the GA structure and the other on the GB
structure (see Supporting Information for details on the modeling procedure). This is a much more
realistic test of the confinement method's ability to accurately calculate relative free energies.

@@

We decomposed the calculated free energy into per residue decomposition as described in method section. This help us for
identification of important residues that stabilizes a particular conformation. 
In this direction, We first calculate the per residue free energy for the GA95 and GB95 sequences. Similar to the total 

@@

the fold, where as residues 1-8 and 52-56 are unstructured in $3 \alpha$, but form $\beta$ strand in $4 \beta + \alpha$ structure.
Most of the amino acid residues in the region 1-9 favors $4 \beta + \alpha$ structure and can also seen from the plot 

@@

Leu-7 favors $4 \beta + \alpha$ structure as these hydrophobic residues are oriented towards the protein hydrophobic 

@@

it more stability compare to the $4 \beta + \alpha$ structure (Figure~\ref{fig:orban_full}(B)). The peak 

@@

exposed to the solvent in $4 \beta + \alpha$ conformation as shown in (Figure~\ref{fig:orban_full}(E)), but only partially exposed 

>>>>

\Alberto{ This is benchmarking, good internal, but not relevant. Should go to SI,out, or summarized
    in one sentence:  "After reproducing the results of Cecchini and coworkers we were interested in
the ability to apply this method to larger systems"}

\Alberto{As a first step, we performed several control experiments to verify that our implementation of the
confinement method produces results compatible with previous calculations reported in the
literature.

The method has previously been applied to a 16 amino acid residue $\beta$-hairpin from
protein G, known as BHP \cite{Cecchini2009}. We calculated the free energy difference between two different
conformations of the peptide: (1) the native conformation, called bhp1, with a two-stranded
$\beta$-sheet; and (2) a conformation, called bhp3, which has a three-stranded $\beta$-sheet.
Analysis of long (4 $\mu$s) equilibrium simulations \cite{Cecchini2009},\cite{Krivov2004}  shows that bhp1 is the more favorable
configuration by 1.8 kcal/mol. Using the confinement method, we obtain a value 1.7 kcal/mol, which
is in good agreement with the equilibrium simulations and with previous calculations using the
confinement method \cite{Cecchini2009} (see Supporting Information for further details).}

Previous applications of the confiment method have focused on relatively short peptides, up to $17$
residues in length. In this work we apply the confinement method to
larger proteins. On that direction, we first test the confinement method for a chamelon sequence and 
found that it can correctly pick out the prefential structure of the chamelon sequence. Other 
applications we explore in the present work is the re-scoring, or
metaprediction, of structures submitted during the Critical Assessment of Structure Prediction
(CASP) experiment (described in detail later). We computed the relative free energies of predictions
submitted during CASP9, with the expectation that the most native-like prediction will have
the lowest free energy. As a control, for several targets we also calculated the relative free
energy of the experimentally determined structure. \Alberto{should not start with results yet, first
we are describing what we are going to show them} As
Table~\ref{table:casp_control} shows, in most cases, the confinement method correctly assigns a
lower free energy to the experimentally determined structure than to any of the decoys. 
Most interestingly, we decomposed the free energy into its per residue component, which help 
us to identify the residues which stabilize or destabilize a particular conformation of the protein.
%Although differentiating between experimentally determined structures and computer generated predictions is
%not a stringent test of a scoring method---for example, examining the residue-residue packing can
%easily distinguish between the two [ref Rosetta Holes]---the results none the less serve as a useful
%"sanity check".

@@

In general, proteins with similar sequences tend to have similar structures. This idea is the basis
of comparative modeling and fold recognition in protein structure prediction. There are, however,
examples---often referred to as chameleon sequences---of proteins with similar sequences that have
remarkably different structures. Orban and co-workers have designed a sequence of 56-residues that
is marginally stable in one of two possible folds. By mutating key residues in this sequence they
are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). Sequences that adopt a mixed alpha/beta structure similar to Protein G
are denoted as ``GB'', while sequences that form a three-helix bundle are denoted as ``GA''. One
pair of sequences (GA88/GB88) are 88 percent identical in sequence and differ in seven positions. Another pair
(GA95/GB95) are 95 percent identical and differ in three positions. The last pair (GA98/GB98) differ
only in a single tyrosine to alanine mutation.

@@

To avoid this potential problem, we instead computed the relative free energy of two different
structural models for each sequence. One model is based on the GA structure and the other on the GB
structure (see Supporting Information for details on the modeling procedure). This is a much more
realistic test of the confinement method's ability to accurately calculate relative free energies.

@@

We decomposed the calculated free energy into per residue decomposition as described in the methods
section. This helps us to
identify important residues that stabilize a particular conformation. 
\Alberto{This is not Reader Friendly at all. we do not have to describe every little detail. I would
    rather have a table where we have the pairing and the reason for favoring one structure or the
    other. In the main text I would say that the main reason for preferences of one fold or another
    has to do with the relative solvent exposure of the residue in one conformation or another, or
    the possibility of stablishing hydrogen/salt bridges in one conformation. This would be my
preference, ditching everything below until the next section. What do you think?}
In this direction, we first calculate the per residue free energy for the GA95 and GB95 sequences. Similar to the total 

@@

the folds, where as residues 1-8 and 52-56 are unstructured in $3 \alpha$, but form $\beta$ strand in $4 \beta + \alpha$ structure.
Most of the amino acid residues in the region 1-9 favor $4 \beta + \alpha$ structure as can be seen from the plot 

@@

Leu-7 favor $4 \beta + \alpha$ structure as these hydrophobic residues are oriented towards the protein hydrophobic 

@@

it more stability compared to the $4 \beta + \alpha$ structure (Figure~\ref{fig:orban_full}(B)). The peak 

@@

exposed to the solvent in $4 \beta + \alpha$ conformation as shown in
(Figure~\ref{fig:orban_full}(E)), but is only partially exposed 
